---
title: "Association between ambient black carbon and temperature and High-density lipoprotein"
author: "Yuhong Hu"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---
<br>
Link to my pdf: [report](https://github.com/rainane007/pm566_finalproject/blob/main/report/report.pdf)       [[download]](https://github.com/rainane007/pm566_finalproject/raw/main/report/report.pdf)
<br>
<br>


```{r setup, message=FALSE, echo=FALSE, warning=FALSE}

library(dplyr)
library(ggplot2)
library(ggpubr)
library(tidyr)
library(data.table)
library(lubridate)
library(skimr)
library(table1)
library(kableExtra)
library(plotly)
library(DT)
library(knitr)

# INITALIZE CODE CHUNK OPTIONS
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = FALSE,
  cache = FALSE,
  fig.width = 8, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px",
  class.source = "code-r")
```

# Introduction
<br>

## Background

High-density lipoprotein (HDL) cholesterol is known as the 'good' cholesterol because it helps remove other forms of cholesterol from bloodstream. Higher levels of HDL cholesterol are associated with a lower risk of heart disease.

Air pollution is known to be one of the leading causes of cardiovascular disease. Emerging evidence suggests that particulate-mediated HDL dysfunction might be a novel mechanism linking air pollution exposure to adverse cardiovascular effects.However, few studies have evaluated the impact of traffic-related air pollution exposure (black carbon, nitrogen dioxide). Similarly, temperature has been linked to cardiovascular disease, but little is known about the underlying mechanisms.
<br>

## Research objectives and questions

Given the background, the project investigated the association between acute exposure to ambient black carbon (one source of traffic-related air pollution) and ambient temperature and HDL level. Additionally, to account for potential latency in HDL level changes, the project also assessed the association between 1-day and 2-day lag exposure and HDL level. Furthermore, the project also explored the effect modification by potential risk factors.

By visualization and modeling, **the project aimed to address the following questions**.

1. Is daily ambient black carbon associated with HDL level? Is there lag effect for the association of interest?

2. Is daily ambient temperature associated with HDL level? Is there lag effect for the association of interest?

3. What are potential 'risk factors' of HDL changes among study population?

4. Does these risk factors modify the association between daily ambient black carbon/temperature and HDL level?

<br>

# Method
<br>

## Dataset and study population

For data disclosure issue,the project was based on part of the Veterans Administration Normative Aging Study (NAS), which could be used for class. NAS was  a longitudinal study established in 1963, the study enrolled 2,280 men from the Greater Boston area, who were aged between 21 to 80 and were determined to be free of known chronic medical conditions by an initial health screening. Participants visited the study center repeatedly for physical examinations, blood pressure measurements, blood sample collection, and questionnaires approximately every four years. Blood samples were used for lipid analysis.

During the follow-up period, black carbon (BC) was measured at a central monitoring site located on the roof of Countway Library, Harvard Medical School, in Boston, MA, temperature was also collected. Single day lags were computed for these pollutants and meteorological variables from the same day of health visits and up to 2 days before the visit. 

```{r EDA,results='hide'}
#load data
nas_hdl <- readRDS("data/nas_hdl.rds")

# Look at the dimension of the dataset and check the missing values/implausible values
str(nas_hdl)
length(unique(nas_hdl$ID))
skim(nas_hdl)

#Exclude missing values and double check the final dataset
nas <- na.omit(nas_hdl)
length(unique(nas$ID))
skim(nas)
summary(nas$YEAR)
```

The dataset included a subset of 981 subjects with a total of 2483 observations. As NAS dataset was well-curated, so the dataset was pretty clean and less subject to missing data issue. For total of 31 variable, `RACE` had 29 missing observations, and `NEDUC` had 1 missing value. By looking at the summary statistics of each variable, we did not observe any implausible value for continuous variable or wrongly-coded value for categorical variable. Therefore, we simply excluded the observations with missing observations, as the missing data issue here is trivial. The final main dataset had 968 subjects with a total of 2453 observations from 1995 to 2011.
<br>

## Data wrangling and transformation

- We first create a new categorical variable `hdlcat` to divide the HDL level into 2categories based on clinical recommendation. For easy plotting and modeling, we also level and label the categorical variable. For tabling summary statistics, we created label for key variables.

```{r wrangling,results='hide'}
# create a vis variable to indicate the visit sequence for each subject for plotting and descriptive statistics
# nas <- nas %>% 
#   arrange(ID,VISIT) %>% 
#   group_by(ID) %>% 
#   mutate(vis = row_number()) %>% 
#   ungroup()

# Level and label the categorical variable for plotting
quantile(nas$AGE)
nas <- nas %>% 
  mutate(hdlcat = factor(case_when(HDL>=40~2,HDL<40~1),
                         levels = c(1,2),
                         labels=c('At risk (<40)','Normal (>=40)')),
         bmicat = factor(case_when(BMI>30~4,(BMI>25&BMI<=30)~2,(BMI>18.5&BMI<=25)~2,BMI<=18.5~1),
                         levels = c(1,2,3,4),
                         labels=c('Underweight','Normal','Overweight','Obese')),
         agecat = factor(case_when(AGE>=73~2,AGE<73~1),
                         levels = c(1,2),
                         labels=c('<73 yrs','>=73 yrs')))

nas <- nas %>% 
  mutate(WARMCOLD = factor(WARMCOLD,levels = c(0,1),labels = c('Cold','Warm')),
         DIABETE = factor(DIABETE,levels = c(0,1),labels = c('No','Yes')),
         STATIN = factor(STATIN,levels = c(0,1),labels = c('No','Yes')),
         HT = factor(HT,levels = c(0,1),labels = c('No','Yes')),
         RACE = factor(RACE,levels = c(1,2,3,4,5),labels = c('White','Black','Hispanic White','Hispanic Black','American Indian')),
         SMK = factor(SMK,levels = c(1,3,4),labels = c('Never','Current','Former')))

# Label variable
label(nas$WARMCOLD) <- 'SEASON'
label(nas$STATIN) <- 'STATIN USE'
label(nas$SMK) <- 'SMOKING STATUS'
label(nas$BC24H) <- 'SAME DAY BLACK CARBON (ug/m3)'
label(nas$TEMPC24H) <- 'SAME DAY TEMPERATURE (°C)'
label(nas$AGE) <- 'AGE (years)'
label(nas$BMI) <- 'BMI (kg/m2)'
label(nas$bmicat) <- 'BMI CATEGORY'
```

- In one of the result section, we would like to show the univariate association of interest by black carbon/temperature on the same day of each visit (`lag0`), black carbon/temperature on the previous 1 day of each visit (`lag1`), black carbon/temperature on the previous 2 days of each visit (`lag2`). However, the lag exposure measures were storaged in 'wide' format, thus, we transformed the original dataset (one subject per row) into 'long' dataset (one time per row, each subject may have muptiple rows) for easy plotting.

- The ambient black carbon and temperature exposure were the same for subjects who took hospital vist on the same day, thus, we had duplicates of the two exposure in the dataset. To plot time series of temperature and black carbon during following, we created dataset without duplicates for variables `BC24H` and `TEMPC24H`.

- Last but not least, the dataset produced without further usage was removed to keep environment tidy.
<br>
```{r transformation}
# Create long format dataset
nas_long1 <- gather(nas, lag, BC, BC24H:BCD2)

nas_long1 <- nas_long1 %>% 
  mutate(lag = case_when(lag=='BC24H'~'lag0',lag=='BCD1'~'lag1',lag=='BCD2'~'lag2'))

nas_long2 <- gather(nas, lag, Temp, TEMPC24H:TMPCD2)

nas_long2 <- nas_long2 %>% 
  mutate(lag = case_when(lag=='TEMPC24H'~'lag0',lag=='TMPCD1'~'lag1',lag=='TMPCD2'~'lag2'))

nas_long <- merge(nas_long1,nas_long2,sort=TRUE)


# Create dataset without duplicate exposure
nas$date <- as.Date(nas$DATE,origin='1960-01-01')

nas_BC <- nas %>% 
  distinct(date,BC24H)

nas_TM <- nas %>% 
  distinct(date,TEMPC24H)

# clean the unnecesary dataset
rm(nas_long1)
rm(nas_long2)
```

## Tools for data exploration

- The packages used for EDA: skimr.

- The packages used for data wrangling and transformation: dplyr, tidyr.

- The packages used for visulization: ggplot2, plotly, ggpubr, table1, kableExtra.

<br>

# Results
<br>

## Temporal trends of daily HDL, daily ambient black carbon and daily ambient temperature during follow-up

- We did not observe a seasonal pattern of population HDL level, but the smooth curve suggested that the HDL level among the participants increased from 1995 to 2000.

- In general, the ambient black carbon level has decrease during follow-up, from 1995 to 2005.

- Regular seasonal fluctuation was observed for ambient temperature during follow-up, as expected. No long-term trend of ambient temperature was observed.

```{r timeseries}
# nas %>% 
#   group_by(date) %>% 
#   summarise(HDL=mean(HDL)) %>% 
#   ggplot(aes(x=date, y=HDL))+
#   geom_line(aes(x=date, y=HDL),color='#9999CC')+
#   labs(title = "Temporal trends of daily average HDL level", x  = "Date", y = "HDL, high-density lipoproteins (mg/dL)")+
#   theme_bw() +
#   theme(
#     plot.title = element_text(face = "bold", size = 12))

trendhdl <- nas %>% 
  ggplot(aes(x=date, y=HDL))+
  geom_smooth(color='#9999CC')+
  geom_point(color='#9999CC',alpha=0.4,size=1)+
  labs(title = "Temporal trends of daily HDL level", x  = "Date", y = "HDL, high-density lipoproteins (mg/dL)")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))

trendbc <- nas_BC %>% 
  ggplot(aes(x=date, y=BC24H))+
  geom_point(color='#0066CC',alpha=0.4,size=1)+
  geom_smooth(color='#0066CC')+
  labs(title = "Temporal trends of daily black carbon", x  = "Date", y = "Black carbon (ug/m3)")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))

trendtemp <- nas_TM %>% 
  ggplot(aes(x=date, y=TEMPC24H))+
  geom_point(color="#66CC99",alpha=0.4,size=1)+
  geom_smooth(color="#66CC99")+
  labs(title = "Temporal trends of daily temperature", x  = "Date", y = "Temperature (°C)")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))

```

### {.tabset}

#### HDL
```{r}
ggplotly(trendhdl)
```

#### Black carbon
```{r}
ggplotly(trendbc)
```

#### Temperature
```{r}
ggplotly(trendtemp)
```


### {-}

## Descriptive statistics

According to clinical cutoff, males with less than 40 mg/dL HDL was at risk of cardiovascular disease. The table showed the basic summary statistics of key variables by the two levels of HDL.

Participants of normal HDL had lower `BMI`, lower daily `black carbon`, and lower daily `temperature`, compared to participants with HDL below the cutoff. The proportion of `obeses` was higher in participants with HDL at risk. The proportion of `diabetes` was lower in participants of normal HDL. During the `cold season`, the HDL levels was more likely to be above the cutoff than during the `warm season`.

```{r table1}
table1(~AGE+BMI+bmicat+RACE+STATIN+DIABETE+WARMCOLD+SMK+BC24H+TEMPC24H|hdlcat,data=nas)%>% 
  kable(caption = "Summary Statistics of (repeated) measurement of characterstics by HDL level") %>% 
  kable_styling()
```

## Distribution of HDL by different levels of risk factors

To further explore whether some demographic and physiological variables were associated with HDL levels, we displayed violin plot (categorical variables) and scatterplot (continuous variables) to show the distribution of HDL by different levels of potential risk factors, and to identify risk factors of HDL among NAS population.

- For categorical variables, the different shapes of violin were observed for different levels of `STATIN`, `DIABETE`, `RACE`. 

- For continuous variables, `AGE` was slightly positively associated with HDL (r = 0.092, p-value < 0.05), and `BMI` was moderately negatively associated with HDL (r = -0.27, p-value < 0.05).

```{r riskfactor}
#risk factor of HDL-statin
statin <- nas %>%
  ggplot(aes(y=HDL,x=STATIN,fill=STATIN))+
  geom_violin(alpha=0.4)+
  geom_boxplot(width=0.1, fill="white",color='grey')+
  labs(title="Distribution of HDL by statin use",
       x = "HDL, high-density lipoproteins (mg/dL)",
       fill = "Statin use")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    legend.position="bottom")+
  scale_fill_brewer(palette="Spectral")


#risk factor of HDL-diabetes
diabetes <- nas %>%
  ggplot(aes(y=HDL,x=DIABETE,fill=DIABETE))+
  geom_violin(alpha=0.4)+
  geom_boxplot(width=0.1, fill="white",color='grey')+
  labs(title="Distribution of HDL by diabetes status",
       x = "HDL, high-density lipoproteins (mg/dL)",
       fill = "Diabetes")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    legend.position="bottom")+
  scale_fill_manual(values=c("#CC6666", "#9999CC"))


#risk factor of HDL-race
race <- nas %>%
  ggplot(aes(y=HDL,x=RACE,fill=RACE))+
  geom_violin(alpha=0.4)+
  geom_boxplot(width=0.1, fill="white",color='grey')+
  labs(title="Distribution of HDL by Race",
       x = "HDL, high-density lipoproteins (mg/dL)",
       fill = "RACE")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    legend.position="bottom")+
  scale_fill_brewer(palette="Spectral")


#risk factor of HDL-season
season <- nas %>%
  ggplot(aes(y=HDL,x=WARMCOLD,fill=WARMCOLD))+
  geom_violin(alpha=0.4)+
  geom_boxplot(width=0.1, fill="white",color='grey')+
  labs(title="Distribution of HDL by warm/cold seasons",
       x = "HDL, high-density lipoproteins (mg/dL)",
       fill = "Season")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    legend.position="bottom")+
  scale_fill_brewer(palette="Spectral")

#risk factor of HDL-SMK
smk <- nas %>%
  ggplot(aes(y=HDL,x = SMK,fill=SMK))+
  geom_violin(alpha=0.4)+
  geom_boxplot(width=0.1, fill="white",color='grey')+
  labs(title="Distribution of HDL by smoking status",
       x = "HDL, high-density lipoproteins (mg/dL)",
       fill = "Smoking")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    legend.position="bottom")+
  scale_fill_manual(values=c("#CC6666", "#9999CC", "#66CC99"))

#risk factor of HDL-HT
ht <- nas %>%
  ggplot(aes(y=HDL,x =HT,fill=HT))+
  geom_violin(alpha=0.4)+
  geom_boxplot(width=0.1, fill="white",color='grey')+
  labs(title="Distribution of HDL by hypertension status",
       x = "HDL, high-density lipoproteins (mg/dL)",
       fill = "Hypertension")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    legend.position="bottom")+
  scale_fill_brewer(palette="Spectral")


#risk factor of HDL-dow
dow <- nas %>%
  ggplot(aes(y=HDL,x=WKDAY,fill=WKDAY))+
  geom_violin(alpha=0.4)+
  geom_boxplot(width=0.1, fill="white",color='grey')+
  labs(title="Distribution of HDL by days of the week",
       x = "HDL, high-density lipoproteins (mg/dL)",
       fill = "Days of the week")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    legend.position="bottom")+
  scale_fill_brewer(palette="Spectral")


#risk factor of HDL-age
age <- nas %>%
  ggplot(aes(x=AGE,y=HDL))+
  geom_point(size=0.5,alpha=0.3,color='#66CC99')+
  geom_smooth(method = lm,color='#336633')+
  # stat_cor(method = "pearson", label.x = 80, label.y = 15)+
  labs(title="Correlation between HDL and age",
        x ="Age (years)  ", y = "HDL, high-density lipoproteins (mg/dL)") +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))


#risk factor of HDL-BMI
bmi <- nas %>%
  ggplot(aes(x=BMI,y=HDL))+
  geom_point(size=0.5,alpha=0.3,color='#CC6666')+
  geom_smooth(method = lm,color='#D55E00')+
  # stat_cor(method = "pearson", label.x = 40, label.y = 15)+
  labs(title="Correlation between HDL and BMI",
        x ="BMI (kg/m2)  ", y = "HDL, high-density lipoproteins (mg/dL)") +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))
```

### {.tabset}

#### Statin
```{r}
ggplotly(statin)
```

#### Diabetes
```{r}
ggplotly(diabetes)
```

#### Race
```{r}
ggplotly(race)
```

#### Season
```{r}
ggplotly(season)
```

#### Smoking status
```{r}
ggplotly(smk)
```

#### Hypertension
```{r}
ggplotly(ht)
```

#### Days of the week
```{r}
ggplotly(dow)
```

#### Age
```{r}
ggplotly(age)
```

#### BMI
```{r}
ggplotly(bmi)
```

### {-}


## Univariate relationship between daily (lag) ambient black carbon, temperature and HDL

Before plotting, we checked the distribution of HDL and found right skewness.The normality was achieved by log transformation of HDL.Therefore, for the following analyses, we used log(HDL) as outcome measure.

```{r distribution}
distribution1 <- nas %>% 
  ggplot() + 
  geom_histogram(aes(x=HDL, y=..density..), fill="#66CC99", alpha=0.4,bins=20) + 
  geom_density(aes(x=HDL),color = '#336633',alpha=0.4)+
  labs(title = "Distribution of HDL", x  = "HDL (mg/dL)")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))

distribution2 <-nas %>% 
  ggplot() + 
  geom_histogram(aes(x=log(HDL), y=..density..), fill="orange", alpha=0.4,bins=20) + 
  geom_density(aes(x=log(HDL)),color = '#D55E00',alpha=0.4)+
  labs(title = "Distribution of log(HDL)", x  = "log(HDL, in mg/dL)")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))

```

### {.tabset}

#### HDL

```{r}
ggplotly(distribution1)
```

#### Log(HDL)
```{r}
ggplotly(distribution2)
```

### {-}

### Effect and lag effect of ambient black carbon and modification by risk factors {.tabset}

```{r relationship}
# Relationship between black carbon and HDL for each lag
bc <- nas_long %>% 
  ggplot()+
  # geom_point(aes(x=BC,y=log(HDL),color=lag),size=0.2,alpha=0.6)+
  geom_smooth(aes(x=BC,y=log(HDL),color=lag),size = 0.5)+
  facet_wrap(~lag)+
  labs(title = "Relationship between daily black carbon and log(HDL) for each lag ", x  = "Black carbon (ug/m3)", y = "log(HDL, in mg/dL)",color='lag')+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))

bcdia <- nas_long %>% 
  ggplot()+
  geom_smooth(aes(x=BC,y=log(HDL),color=DIABETE,fill=DIABETE),size = 0.5)+
  facet_wrap(~lag)+
  labs(title = "Relationship between daily black carbon and log(HDL) for by diabaetes each lag ", x  = "Black carbon (ug/m3)", y = "log(HDL, in mg/dL)")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))

bcsta <- nas_long %>% 
  ggplot()+
  geom_smooth(aes(x=BC,y=log(HDL),color=STATIN,fill=STATIN),size = 0.5)+
  facet_wrap(~lag)+
  labs(title = "Relationship between daily black carbon and log(HDL) by statin use for each lag ", x  = "Black carbon (ug/m3)", y = "log(HDL, in mg/dL)")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))

bcbmi <- 
  nas_long %>% 
  filter(bmicat!='Underweight') %>% ## we do not have enough sample size for underweight
  ggplot()+
  geom_smooth(aes(x=BC,y=log(HDL),color=bmicat,fill=bmicat),size = 0.5)+
  facet_wrap(~lag)+
  labs(title = "Relationship between daily black carbon and log(HDL) by BMI category for each lag ", x  = "Black carbon (ug/m3)", y = "log(HDL, in mg/dL)")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))

bcage <- nas_long %>% 
  ggplot()+
  geom_smooth(aes(x=BC,y=log(HDL),color=agecat,fill=agecat),size = 0.5)+
  facet_wrap(~lag)+
  labs(title = "Relationship between daily black carbon and log(HDL) by median age for each lag ", x  = "Black carbon (ug/m3)", y = "log(HDL, in mg/dL)")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))

```

- The pattern of associations were similar at lag0 and lag1, which were different at lag2.

- In terms of overall association, same day black carbon (lag0) , previous day black carbon (lag1), and previous two days black carbon (lag2) seemed to inversely and linearly associated with HDL level. The magnitude of the association was similar across the lags.

- Effect heterogeneity of diabetes was observed for the association of interest at lag2, where null association found for diabetes.

- Among people with greater age, the association of interest was stronger than people with age below median at lag0 and lag2.

- No clear pattern were shown for modification by statin use and BMI.

#### Overall

```{r}
ggplotly(bc)
```

#### Diabetes

```{r}
ggplotly(bcdia)
```

#### Statin Use

```{r}
ggplotly(bcsta)

```

#### BMI

```{r}
ggplotly(bcbmi)
```

#### Age

```{r}
ggplotly(bcage)
```


### {-}

### Effect and lag effect of temperature and modification by risk factors {.tabset}

```{r}
# Relationship between temperature and HDL for each lag
temp <- nas_long %>% 
  ggplot()+
  # geom_point(aes(x=Temp,y=log(HDL),color=lag),size=0.2,alpha=0.6)+
  geom_smooth(aes(x=Temp,y=log(HDL),color=lag),size = 0.5)+
  facet_wrap(~lag)+
  labs(title = "Relationship between daily temperature and log(HDL) for each lag ", x  = "Temperature (°C)", y = "log(HDL, in mg/dL)")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))


tempdia <- nas_long %>% 
  ggplot()+
  # geom_point(aes(x=Temp,y=log(HDL),color=lag),size=0.2,alpha=0.6)+
  geom_smooth(aes(x=Temp,y=log(HDL),color=DIABETE,fill=DIABETE),size = 0.5)+
  facet_wrap(~lag)+
  labs(title = "Relationship between daily temperature and log(HDL) by diabetes for each lag ", x  = "Temperature (°C)", y = "log(HDL, in mg/dL)")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))

tempsta<- nas_long %>% 
  ggplot()+
  # geom_point(aes(x=Temp,y=log(HDL),color=lag),size=0.2,alpha=0.6)+
  geom_smooth(aes(x=Temp,y=log(HDL),color=STATIN,fill=STATIN),size = 0.5)+
  facet_wrap(~lag)+
  labs(title = "Relationship between daily temperature and log(HDL) by statin use for each lag ", x  = "Temperature (°C)", y = "log(HDL, in mg/dL)")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))

tempbmi<- nas_long %>% 
  filter(bmicat!='Underweight') %>% 
  ggplot()+
  # geom_point(aes(x=Temp,y=log(HDL),color=lag),size=0.2,alpha=0.6)+
  geom_smooth(aes(x=Temp,y=log(HDL),color=bmicat,fill=bmicat),size = 0.5)+
  facet_wrap(~lag)+
  labs(title = "Relationship between daily temperature and log(HDL) by BMI for each lag ", x  = "Temperature (°C)", y = "log(HDL, in mg/dL)")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))


tempage<- nas_long %>% 
  ggplot()+
  # geom_point(aes(x=Temp,y=log(HDL),color=lag),size=0.2,alpha=0.6)+
  geom_smooth(aes(x=Temp,y=log(HDL),color=agecat,fill=agecat),size = 0.5)+
  facet_wrap(~lag)+
  labs(title = "Relationship between daily temperature and log(HDL) by median age for each lag ", x  = "Temperature (°C)", y = "log(HDL, in mg/dL)")+
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 12))

```

- The pattern of associations were similar across three lags.

- In terms of overall association, the relationship was inverted U-shape, with average HDL increasing below around 10 °C and decreasing between temperatures 10 and 30 °C.

- Effect heterogeneity of diabetes was observed, where null association was found for diabetes.

- For people using statin use, there was an inverse association between HDL and temperature. However, among people without statin usage, the relationship was inverted U-shape, with average HDL increasing below around 10 °C and decreasing between temperatures 10 and 30 °C.

- Among people with greater age, there was an inverse association between HDL and temperature. However, among people with age lower than median, the relationship was inverted U-shape, with average HDL increasing below around 10 °C and decreasing between temperatures 10 and 30 °C.

- No clear pattern were shown for BMI.

#### Overall

```{r}
ggplotly(temp)
```

#### Diabetes

```{r}
ggplotly(tempdia)
```

#### Statin Use

```{r}
ggplotly(tempsta)
```

#### BMI

```{r}
ggplotly(tempbmi)
```

#### Age

```{r}
ggplotly(tempage)
```


### {-}

# Conclusion

The relationship between same-day and lagged temperature and HDL levels was inverted U-shape, with average HDL increasing below around 10 °C and decreasing between temperatures 10 and 30 °C. In contrast, we found inverse and linear association between same-day and lagged exposure of ambient black carbon and HDL levels, this was further supported by the result of linear mixed effect model where we did detect a significant effect of same-day black carbon and its two lags on HDL (See Appendix). Furthermore, diabetes and age seemed to modify the association between daily temperature and HDL levels across all the lags assessed. Participants with diabetes showed null association between daily temperature and HDL levels whereas non-diabetes showed U-shaped similar to the main effect pattern. Participants above the median age showed that HDL levels decreased as the temperature increased, but in those below the median age, similar U-shaped was observed.

